---
title: "MATH163 – Week 07: TRY SOLUTIONS"
output: html_notebook
---

```{r}
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(here))
set.seed(163)
```

# A. Load & tidy
```{r message=FALSE, warning=FALSE}
excel_path <- here::here("data", "exchangerates.xlsx")
stopifnot(file.exists(excel_path))
exrates <- readxl::read_excel(excel_path, skip = 5)

UK_col    <- grep("^United Kingdom$", names(exrates))
China_col <- grep("^China \\(People's Republic of\\)$", names(exrates))
UK    <- as.numeric(exrates[[UK_col]])
China <- as.numeric(exrates[[China_col]])
UK    <- UK[!is.na(UK)]
China <- China[!is.na(China)]
summary(UK); summary(China)
```

# [7-TY-1] China analysis (n=5 and n=50)
```{r}
B <- 500
n5  <- 5
n50 <- 50

CH_means_5  <- replicate(B, mean(sample(China, size = n5,  replace = TRUE)))
CH_means_50 <- replicate(B, mean(sample(China, size = n50, replace = TRUE)))

par(mfrow = c(1,2))
hist(CH_means_5,  freq = FALSE, main = "China means (n=5)",  xlab = "mean", col = "lightgray")
curve(dnorm(x, mean = mean(China), sd = sd(China)/sqrt(n5)), add = TRUE, lwd = 2)
hist(CH_means_50, freq = FALSE, main = "China means (n=50)", xlab = "mean", col = "lightgray")
curve(dnorm(x, mean = mean(China), sd = sd(China)/sqrt(n50)), add = TRUE, lwd = 2)
par(mfrow = c(1,1))
```

**Comment:** At the same `n`, whichever country has larger population sd will show wider means. Increasing `n` tightens the distribution (CLT).

# [7-TY-2] UK means for n ∈ {5,10,50,100}
```{r}
B <- 500
ns <- c(5,10,50,100)
UK_means_list <- lapply(ns, function(n) replicate(B, mean(sample(UK, n, replace = TRUE))))

par(mfrow = c(2,2))
for (i in seq_along(ns)) {
  n <- ns[i]
  hist(UK_means_list[[i]], freq = FALSE,
       main = paste0("UK means (n=", n, ")"), xlab = "mean", col = "lightgray")
  curve(dnorm(x, mean = mean(UK), sd = sd(UK)/sqrt(n)), add = TRUE, lwd = 2)
}
par(mfrow = c(1,1))
```

**Comment:** Spread clearly shrinks roughly as `1/sqrt(n)`.

# [7-TY-3] Pooled series (UK + China) with Normal overlay
```{r}
both <- c(UK, China)
B <- 500; n <- 50
both_means <- replicate(B, mean(sample(both, size = n, replace = TRUE)))

hist(both_means, freq = FALSE, main = "Means from pooled series (n=50)",
     xlab = "mean", col = "lightgray")
curve(dnorm(x, mean = mean(both), sd = sd(both)/sqrt(n)), add = TRUE, lwd = 2)
```

**Comment:** The Normal overlay is typically very close by `n = 50` (CLT).

# [7-TY-4] Optional: t overlay for n=10
```{r}
n <- 10
UK_means_n10 <- replicate(500, mean(sample(UK, size = n, replace = TRUE)))
hist(UK_means_n10, freq = FALSE, main = "UK means (n=10): Normal vs t", xlab = "mean", col = "lightgray")

# Normal overlay
curve(dnorm(x, mean = mean(UK), sd = sd(UK)/sqrt(n)), add = TRUE, lwd = 2)

# t overlay (approximate visual): center at mean(UK); scale sd(UK)/sqrt(n); df = n-1
xs <- seq(min(UK_means_n10), max(UK_means_n10), length.out = 400)
mu  <- mean(UK); s <- sd(UK)/sqrt(n); df <- n - 1
lines(xs, dt((xs - mu)/s, df = df)/s, lwd = 2, lty = 2)
legend("topright", legend = c("Normal", "t (df=9)"),
       lwd = 2, lty = c(1,2), bty = "n")
```

**Comment:** For moderate `n` the Normal and t overlays are very similar on the means scale. Differences become negligible as `n` grows.
