---
title: "MATH163 – Week 05: Generating Multiple Samples"
output: html_notebook
---

```{r}
set.seed(163)
```

# INTRODUCTION

In this lab we’ll:
- generate many random samples from a known distribution (Geometric),
- store them as columns of a matrix (so each column is one sample),
- compute column summaries (means, medians, minima, maxima),
- and see how these summaries vary from sample to sample.

We’ll also do a quick visual check of the classic approximation
Binomial(n, p) ≈ Poisson(λ = n p) for small p.

---

# WORKED EXAMPLE (WITH ANSWERS)

## 1) Build the same 100×150 dataset three ways

Goal: create a matrix with 100 rows × 150 columns, where each column is a sample of size 100 from a Geometric distribution with parameter p = 0.1.
`rgeom(n, p)` generates counts of failures before the first success; values are 0,1,2,…

```{r}
# Method A: generate one long vector then reshape it
m1 <- rgeom(100 * 150, prob = 0.1)  # 15,000 numbers in one go
dim(m1) <- c(100, 150)              # give it matrix dimensions (100 rows, 150 cols)

# Method B: pre-allocate a matrix and fill row-by-row in a for-loop
m2 <- matrix(NA_integer_, nrow = 100, ncol = 150)
for (i in 1:100) {
  # each loop fills the i-th row with 150 draws
  m2[i, ] <- rgeom(150, prob = 0.1)
}

# Method C: use replicate() to make 150 columns, each column a length-100 draw
m3 <- replicate(150, rgeom(100, prob = 0.1))  # result is 100×150
```

## 2) Compare sample means and sample medians across columns

```{r}
# Column-wise summaries
means_m1   <- colMeans(m1)                # fast mean per column
medians_m1 <- apply(m1, 2, median)        # general "apply over columns"

# Side-by-side histograms to compare their distributions
op <- par(mfrow = c(1, 2))  # 1 row, 2 plots
hist(means_m1,
     main = "100-sample means (Geom p=0.1)",
     xlab = "mean of one column")
hist(medians_m1,
     main = "100-sample medians (Geom p=0.1)",
     xlab = "median of one column")
par(op)  # restore layout
```

## 3) Minima and maxima (shape intuition)

```{r}
mins_m1 <- apply(m1,  2, min)
maxs_m1 <- apply(m1,  2, max)

op <- par(mfrow = c(1, 2))
hist(mins_m1, main = "Column minima (Geom p=0.1, n=100)",
     xlab = "min per column")
hist(maxs_m1, main = "Column maxima (Geom p=0.1, n=100)",
     xlab = "max per column")
par(op)
```

## 4) Binomial → Poisson overlay (one case)

We’ll simulate a Binomial with n = 100, p = 0.05 many times, approximate its PMF by relative frequencies, and overlay the PMF of Poisson with λ = n p = 5.

```{r}
# Empirical Binomial PMF via many draws
n   <- 100
p   <- 0.05
lam <- n * p

B   <- rbinom(2e5, size = n, prob = p)    # many draws for a smooth empirical PMF
tab <- table(B)
k   <- as.integer(names(tab))             # support values with at least one count
emp <- as.numeric(tab) / sum(tab)         # empirical probability mass at each k

# Theoretical Poisson PMF at the same support
pois_pmf <- dpois(k, lambda = lam)

# Plot empirical PMF as bars and overlay the Poisson curve
plot(k, emp, type = "h", lwd = 3,
     main = "Binomial(n=100,p=0.05) vs Poisson(λ=5)",
     xlab = "k", ylab = "Probability")
lines(k, pois_pmf, lwd = 2)
legend("topright", legend = c("Empirical Binomial", "Poisson(λ=5)"),
       lwd = c(3,2), bty = "n")
```

---

# TRY IT YOURSELF (NO ANSWERS HERE)

- **[5-TY-1]** Build a 50×300 matrix of Geom(p = 0.2) using `replicate()`.
  Make side-by-side histograms of column means and medians; comment on how they compare.

- **[5-TY-2]** Using `m1`, draw histograms of column minima and column maxima.
  Explain the shapes in terms of the geometric distribution and sample size per column.

- **[5-TY-3]** For n = 100, run the Binomial→Poisson visual check for p ∈ {0.01, 0.05, 0.1}.
  For each p:
  1) generate many Binomial draws,
  2) barplot the empirical PMF,
  3) overlay Poisson(λ = n p).
  A visual assessment is fine—note when/why the approximation looks best.
