---
title: "MATH163 – Week 07: Sampling from a Real Dataset"
output: html_notebook
---

```{r}
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(here))
set.seed(163)
```

# INTRODUCTION

In this lab you will:

- Import monthly **exchange rates** from Excel (OECD file) for the UK and China.
- Treat each country's column as a **population** for the purpose of simulation.
- Draw many random **samples with replacement**.
- Look at the distributions of **sample means** and **sample variances** and how they change with the sample size `n`.
- See a visual form of the **Central Limit Theorem** by overlaying a Normal curve on the histogram of sample means.

**Data location:** put `exchangerates.xlsx` in a `data/` subfolder under your course folder (same location you used in previous weeks).

```{r message=FALSE, warning=FALSE}
# --- LOAD DATA (with a brief check) ----------------------------------
excel_path <- here::here("data", "exchangerates.xlsx")
if (!file.exists(excel_path)) {
  stop("I can't find the file at: ", excel_path,
       "\nFix: place 'exchangerates.xlsx' in a 'data' folder under your course folder.")
}
# The first 5 rows are source notes/headers in the OECD sheet, so we skip 5.
exrates <- readxl::read_excel(excel_path, skip = 5)

# Extract the two columns we need robustly (avoid quoting issues in names)
UK_col    <- grep("^United Kingdom$", names(exrates))
China_col <- grep("^China \\(People's Republic of\\)$", names(exrates))
UK    <- as.numeric(exrates[[UK_col]])
China <- as.numeric(exrates[[China_col]])

# Drop NAs that might appear at the tail of the sheet
UK    <- UK[!is.na(UK)]
China <- China[!is.na(China)]

summary(UK); summary(China)
length(UK); length(China)
```

# WORKED EXAMPLE (WITH ANSWERS)

**Goal:** For the UK series, simulate `B = 500` random samples of size `n = 5`, each *with replacement*. For each sample compute the **mean** and **variance**. Then look at the histograms.

```{r}
B <- 500   # number of bootstrap-like samples
n <- 5     # sample size for each draw

UK_means <- replicate(B, mean(sample(UK,   size = n, replace = TRUE)))
UK_vars  <- replicate(B, var( sample(UK,   size = n, replace = TRUE)))

par(mfrow = c(1,2))
hist(UK_means, main = "UK sample means (n=5)", xlab = "mean", col = "grey")
hist(UK_vars,  main = "UK sample variances (n=5)", xlab = "variance", col = "grey")
par(mfrow = c(1,1))
```

**Interpretation:** As `n` increases, the **spread** of the means should **shrink** roughly like \(\sigma/\sqrt{n}\). Variances are more skewed and do not “concentrate” in the same way.

**Quick CLT overlay (Normal) for the means:** We overlay a Normal with mean `mean(UK)` and sd `sd(UK)/sqrt(n)`.

```{r}
par(mfrow = c(1,1))
h <- hist(UK_means, breaks = "FD", main = "UK means (n=5) with Normal overlay",
          xlab = "mean", col = "lightgray", freq = FALSE)
xs <- seq(min(UK_means), max(UK_means), length.out = 400)
curve(dnorm(x, mean = mean(UK), sd = sd(UK)/sqrt(n)), add = TRUE, lwd = 2)
```

# TRY IT YOURSELF (NO ANSWERS HERE)

- **[7-TY-1]** Repeat the UK analysis **for China** (use `n=5` and `n=50`). Compare histograms of the **means** for UK vs China. Which one looks more variable at the same `n`? Why?  
- **[7-TY-2]** For the **UK**, run `n ∈ {5, 10, 50, 100}` with `B = 500` each time. Make a **2×2 grid** of histograms of the sample means. Comment on how the spread changes with `n`.  
- **[7-TY-3]** Combine vectors: `both <- c(UK, China)`. For `n = 50`, draw `B = 500` sample means from `both`. Plot a histogram (density scale) and **overlay a Normal** with mean `mean(both)` and standard deviation `sd(both)/sqrt(n)`. What do you see?  
- **[7-TY-4] (Optional)** Try a **t overlay** for `n = 10` (centered at `mean(UK)`, scale `sd(UK)/sqrt(n)`, df = `n-1`). Does it look different from the Normal overlay at this `n`? (Note: the t-distribution is covered in week 10. This may be familiar from A-level, otherwise you encouraged to look it up in advance.)
